Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-06-01 paper/paper.qmd
Write methods for NNB

To calculate a clinically-relevant NNS for these models, we computed the
confusion matrix at the 95th percentile of risk for each prediction model
(@tbl-risk).
Among the models predicting severe outcomes, those trained on the full
datasets performed best with an NNS of `r allcause_nns` for the all-cause
definition, `r attrib_nns` for the attributable definition, and 
`r pragmatic_nns` for the pragmatic definition.
For context, prior studies predicted CDI-attributable severity using whole
Electronic Health Record data extracted two days after diagnosis and from a
smaller set of manually curated variables, achieving precision values of 0.417
(NNS = `r ehr_nns`) for the EHR model and 0.167 (NNS = `r curated_nns`) for the
curated model at the 95th percentile of risk
[@li_using_2019;@rao_clostridium_2015]. <!-- TODO what about NNS on day of diagnosis? -->
Multiplying the NNS of the OTU-based models by the estimated NNT of 
`r fdx_nnt` for fidaxomicin yields NNB values of `r allcause_fdx_nnb` 
for all-cause severity, `r attrib_fdx_nnb` for attributable severity,
and `r pragmatic_fdx_nnb` for pragmatic severity.
Thus, in a hypothetical scenario where these assumptions about fidaxomicin hold
true, between `r min_fdx_nnb` and `r max_fdx_nnb` patients would need to be
predicted to experience a severe outcome and be treated with fidaxomicin in
order for one patient to benefit.
As the NNS values were computed at the 95th percentile of risk (where 5% of
patients screened are predicted to experience severity), these NNB values mean
that `r min_screen` to `r max_screen` total CDI patients would need to be
screened by an OTU-based prediction model in order for one patient to benefit.
For comparison, pairing the prior EHR-based model with fidaxomicin would yield
an NNB of `r ehr_fdx_nnb` with `r ehr_screen` total CDI patients screened for
one patient to benefit.
These estimates represent a proof-of-concept demonstration of the potential
value of deploying severity prediction models to guide clinicians' treatment
decisions.

@@

@@

::: {#fig-performance fig-cap="**Performance of ML models.**
In the left facets, models were trained on the full datasets, with different
numbers of samples available for each severity definition.
In the right facets, models were trained on the same dataset consisting of the
intersection of samples with labels available for all definitions.
Note that the intersection dataset has exactly the same labels for attributable
and pragmatic severity, thus these have identical performance.
**A)** 
Area under the receiver-operator characteristic curve (AUROC) for the test sets
and cross-validation folds of the training sets, and the area under the balanced
precision-recall curve (AUBPRC) for the test sets.
Each point is the median performance across 100 train/test splits with tails as
the 95% CI.
**B)** Receiver-operator characteristic curves for the test sets.
Mean specificity is reported at each sensitivity value, 
with ribbons as the 95% CI.
**C)** Balanced precision-recall curves for the test sets.
Mean balanced precision is reported at each recall value, 
with ribbons as the 95% CI.
"}
![](figures/ml-performance.png)
:::

>>>>

To calculate a clinically-relevant NNS for these models, we computed the
confusion matrix at the 95th percentile of risk for each prediction model
(@tbl-risk).
We excluded the IDSA severity models as the IDSA severity scores were calculated
on the day of diagnosis, thus they are classification rather than prediction
problems.
Furthermore, IDSA severity scores do not correlate well with disease-related
adverse events which are a more salient outcome to prevent.
Among the models predicting severe outcomes, those trained on the full
datasets performed best with an NNS of `r allcause_nns` for the all-cause
definition, `r attrib_nns` for the attributable definition, and 
`r pragmatic_nns` for the pragmatic definition.
For context, prior studies predicted CDI-attributable severity using whole
Electronic Health Record data extracted two days after diagnosis and from a
smaller set of manually curated variables, achieving precision values of 0.417
(NNS = `r ehr_nns`) for the EHR model and 0.167 (NNS = `r curated_nns`) for the
curated model at the 95th percentile of risk
[@li_using_2019;@rao_clostridium_2015]. <!-- TODO what about NNS on day of diagnosis? -->
Multiplying the NNS of the OTU-based models by the estimated NNT of 
`r fdx_nnt` for fidaxomicin yields NNB values of `r allcause_fdx_nnb` 
for all-cause severity, `r attrib_fdx_nnb` for attributable severity,
and `r pragmatic_fdx_nnb` for pragmatic severity.
Thus, in a hypothetical scenario where these assumptions about fidaxomicin hold
true, between `r min_fdx_nnb` and `r max_fdx_nnb` patients would need to be
predicted to experience a severe outcome and be treated with fidaxomicin in
order for one patient to benefit.
As the NNS values were computed at the 95th percentile of risk (where 5% of
patients screened are predicted to experience severity), these NNB values mean
that `r min_screen` to `r max_screen` total CDI patients would need to be
screened by an OTU-based prediction model in order for one patient to benefit.
For comparison, pairing the prior EHR-based model with fidaxomicin would yield
an NNB of `r ehr_fdx_nnb` with `r ehr_screen` total CDI patients screened for
one patient to benefit.
These estimates represent a proof-of-concept demonstration of the potential
value of deploying severity prediction models to guide clinicians' treatment
decisions.

@@

## Number needed to benefit

For the severity prediction models (which excludes the IDSA definition),
we set out to estimate the potential benefit of deploying models in clincal
settings.
We determined the decision threshold at the 95th percentile of risk for each
model, which corresponds to 5% of cases being predicted by the model to
experience a severe outcome.
At this threshold we computed the number needed to screen (NNS), which is the
reciprocal of precision and represents the number of cases that must be
predicted as severe to identify one true positive [@rembold_number_1998].
The number needed to treat (NNT) is the number of true positive
patients that must be treated by an intervention in order for one patient to
benefit, and is calculated from the reciprocal of absolute risk in randomized
controlled trials [@laupacis_assessment_1988].
Multiplying the NNS of a model by the NNT of a treatment yields the number
needed to benefit (NNB) - the number of patients that must be predicted to have
a severe outcome and undergo a treatment to benefit from it [@liu_number_2019].
NNB encapsulates the benefit of pairing a predictive model with a treatment in
a clinical setting, with lower NNB numbers being better.

@@

::: {#fig-performance fig-cap="**Performance of ML models.**
In the left facets, models were trained on the full datasets, with different
numbers of samples available for each severity definition.
In the right facets, models were trained on the same dataset consisting of the
intersection of samples with labels available for all definitions.
Note that the intersection dataset has exactly the same labels for attributable
and pragmatic severity, thus these have identical performance.
**A)** 
Area under the receiver-operator characteristic curve (AUROC) for the test sets
and cross-validation folds of the training sets, and the area under the balanced
precision-recall curve (AUBPRC) for the test sets.
Each point is annotated with the median performance across 100 train/test splits
with tails as the 95% CI.
**B)** Receiver-operator characteristic curves for the test sets.
Mean specificity is reported at each sensitivity value, 
with ribbons as the 95% CI.
**C)** Balanced precision-recall curves for the test sets.
Mean balanced precision is reported at each recall value, 
with ribbons as the 95% CI.
"}
![](figures/ml-performance.png)
:::
